Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-07 index.qmd
section:introduction,issues,structural; process:write

```{r}
#| results: hide
#| label: code-package

# # load packages
# libraries = c('RColorBrewer','stringr','dplyr','tidyverse',
#               'rechape2','knitr','kableExtra',
#               'rstan','StanHeaders','runjags','rethinking')
# sapply(libraries, require, character.only=T)

```

```{r}
#| results: hide
#| label: code-usd

# # load functions
# main_dir = '/home/josema/Desktop/1. Work/1 research/PhD Antwerp/#thesis/paper2/paper2_manuscript'
# source( file.path( main_dir, 'code', 'user-defined-functions.R') )

```

@@

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level. After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest. This method has been successfully employed in assessing the quality of written texts, where quality describes the underlying trait of interest and the texts serve as the stimuli [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing. The scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020]. A type I error results when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model.  A structural model specify how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s structural and measurement models can offer significant advantages.

@@

From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008.

  What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution

  Table with assumptions

@@

## Appendix A: Ignoring uncertainty {#sec-appendix-A}

```{r}
#| label: code-AppA-simulation

# # simulate units and sub-units
# USd = sim_units_trait( Un=10,
#                        Us=c(1,1),
#                        Ub=c(0,-0.2,0.2),
#                        Useed = 45789,
#                        Sn=6,
#                        Ss=c(0.8,0.8),
#                        Sb=c(0,0,0),
#                        Sseed = 9478 )
# 
# # simulate judges bias
# Jd = sim_judges_bias( Jn=10,
#                       Js=c(0.02,0.02),
#                       Jseed = 79985 )
# 
# 
# # simulate full comparison data
# d = cm_full( USd=USd$USd,
#              Jd=Jd$Jd )
# 
# 
# # set data in list
# dL = list_data( d )
# str(dL)

```

## Appendix B: The five cases of Thurstone {#sec-appendix-B}

<!-- Thurstone's comparative judgment (CJ) theory [@Thurstone_1927] is composed of two key elements: the discriminal process and the law of comparative judgment.   -->

<!-- In a CJ study evaluating text quality quality of written texts -->

>>>>

@@

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence in different stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level. After multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the resulting outcomes and estimate scores that reflect the underlying trait of interest. This method has been successfully employed in assessing the quality of written texts, where quality describes the underlying trait of interest and the texts serve as the stimuli [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests. For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of this bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, p. 25; @Hoyle_et_al_2023, p. 137]. Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020].

<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Kline_et_al_2023; @Hoyle_et_al_2023, p. 138] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

@@

<!-- From early on in the literature, it has been clear that the BTL model represents a statistical articulation of Thurstone's Case 5 [-@Thurstone_1927]. Talk about @Pollitt_et_al_2003 and @Bramley_2008. -->

<!--   What case 5 implies, Assumptions. Not a normal distribution but a logistic distribution -->

<!--   Table with assumptions -->

@@
